---
title: "Getting started with HERITAGE"
author: "Abdoul Oudouss Diakite"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with HERITAGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4,
  fig.align  = "center"
)
set.seed(2026)
```

## What HERITAGE does

`heritage` implements the **HERITAGE** framework (Hierarchical Effects
Regression with Interactions for Trait Analysis in GEnetics). It reinterprets a
pedigree as a directed network of intergenerational influence and decomposes
each phenotype into two parts:

1. a **direct effect** of an individual's own covariates,
   $\boldsymbol{\beta}^\top x_i$; and
2. a **kinship-weighted ancestral spillover** that captures genetic nurture,
   $\boldsymbol{\delta}^\top s_X^{(i)}$.

The conditional mean of the outcome is

$$
\eta_i = \alpha + \boldsymbol{\beta}^\top x_i +
         \boldsymbol{\delta}^\top s_X^{(i)},
\qquad
s_X^{(i)} = \frac{\sum_{j \in \mathrm{Anc}(i)} \phi_{ij}\, x_j}
                 {\sum_{j \in \mathrm{Anc}(i)} \phi_{ij}},
$$

where $\mathrm{Anc}(i)$ is the set of ancestors of $i$ and
$\phi_{ij} = K_{ij}$ is the kinship coefficient. Estimation uses the
Hierarchical Alternating Minimization Algorithm (HAMA), which enforces
**hierarchical sparsity**: a spillover effect $\delta_j$ can be non-zero only
when the corresponding direct effect $\beta_j$ is non-zero,
$\mathrm{supp}(\boldsymbol{\delta}) \subseteq \mathrm{supp}(\boldsymbol{\beta})$.

This vignette walks through a complete, self-contained workflow on a small
simulated cohort. The only required package is `heritage` itself.

```{r library}
library(heritage)
```

## 1. A small pedigree

We build a three-generation pedigree by hand: a set of unrelated founders who
pair up to produce children, who in turn produce grandchildren. Founders have
their `dam` and `sire` set to `0`.

```{r pedigree}
n_founders <- 20L

# Generation 0: founders (unknown parents)
g0 <- data.frame(id = 1:n_founders, dam = 0L, sire = 0L)

# Generation 1: pair consecutive founders -> 2 children per couple
make_children <- function(parents, start_id, n_per_couple = 2L) {
  dams  <- parents$id[seq(1L, nrow(parents), by = 2L)]
  sires <- parents$id[seq(2L, nrow(parents), by = 2L)]
  kids  <- do.call(rbind, Map(function(d, s, k) {
    data.frame(id = k, dam = d, sire = s)
  }, dams, sires,
  split(seq.int(start_id, length.out = length(dams) * n_per_couple),
        rep(seq_along(dams), each = n_per_couple))))
  kids
}

g1 <- make_children(g0, start_id = max(g0$id) + 1L)
g2 <- make_children(g1, start_id = max(g1$id) + 1L)

pedigree <- rbind(g0, g1, g2)
N <- nrow(pedigree)
head(pedigree, 10)
cat(sprintf("Total individuals: %d (founders: %d)\n", N, n_founders))
```

## 2. Genotypes

We simulate `d` independent biallelic markers as 0/1/2 dosages. Founder
genotypes are drawn from Hardy-Weinberg proportions; offspring inherit one
allele from each parent.

```{r genotypes}
d  <- 60L
af <- runif(d, 0.1, 0.9)   # allele frequencies

X <- matrix(0L, N, d)

# Founders: HWE sampling
for (i in seq_len(n_founders)) {
  X[i, ] <- rbinom(d, size = 1, prob = af) + rbinom(d, size = 1, prob = af)
}

# Non-founders: transmit one allele from each parent
draw_allele <- function(g) ifelse(g == 2L, 1L,
                            ifelse(g == 0L, 0L, rbinom(length(g), 1, 0.5)))
for (i in (n_founders + 1L):N) {
  dam_row  <- which(pedigree$id == pedigree$dam[i])
  sire_row <- which(pedigree$id == pedigree$sire[i])
  X[i, ] <- draw_allele(X[dam_row, ]) + draw_allele(X[sire_row, ])
}

dim(X)
```

## 3. Ancestors and the kinship matrix

`build_ancestor_list()` walks the pedigree once and returns, for each
individual, the **row indices** of all of its ancestors. This is exactly the
format `compute_spillover_X()` expects.

```{r ancestors}
ancestors <- build_ancestor_list(pedigree, verbose = FALSE)

ancestors[[1]]                 # a founder -> integer(0)
ancestors[[N]]                 # a grandchild -> parents + grandparents
```

The kinship matrix $K$ holds the kinship coefficient $\phi_{ij}$ for every
pair. In practice you would compute it from pedigree records with
`kinship2::kinship()` or from genotypes with a genomic relationship matrix.
Here we compute it directly with the standard recursive algorithm so the
vignette stays self-contained.

```{r kinship}
kinship_matrix <- function(ped) {
  N <- nrow(ped)
  pos <- match(ped$id, ped$id)
  dam  <- match(ped$dam,  ped$id)   # NA for founders
  sire <- match(ped$sire, ped$id)
  K <- matrix(0, N, N)
  for (i in seq_len(N)) {
    di <- dam[i]; si <- sire[i]
    # diagonal: 0.5 * (1 + kinship(dam, sire))
    K[i, i] <- 0.5 * (1 + if (!is.na(di) && !is.na(si)) K[di, si] else 0)
    if (i > 1L) for (j in seq_len(i - 1L)) {
      kj <- 0
      if (!is.na(di)) kj <- kj + 0.5 * K[di, j]
      if (!is.na(si)) kj <- kj + 0.5 * K[si, j]
      K[i, j] <- K[j, i] <- kj
    }
  }
  K
}

K <- kinship_matrix(pedigree)
K[1:6, 1:6]
```

## 4. Spillover covariates

`compute_spillover_X()` turns genotypes, kinship, and the ancestor list into
the spillover design matrix $s_X$. Founder rows are zero by construction.

```{r spillover}
sX <- compute_spillover_X(X, K, ancestors)

cat(sprintf("Founders with zero spillover: %d / %d\n",
            sum(rowSums(abs(sX)) == 0), n_founders))
```

## 5. Simulate a phenotype

We plant a sparse, hierarchical truth: 5 markers carry a direct effect, and 3
of those *also* carry a spillover effect (a strict subset, as the model
assumes).

```{r outcome}
causal_beta  <- sort(sample(d, 5))
causal_delta <- sort(sample(causal_beta, 3))

beta_true  <- numeric(d); beta_true[causal_beta]   <- rnorm(5, sd = 0.6)
delta_true <- numeric(d); delta_true[causal_delta] <- rnorm(3, sd = 0.4)

y <- 1 + X %*% beta_true + sX %*% delta_true + rnorm(N, sd = 0.5)
y <- as.vector(y)

cat(sprintf("Direct truth   |S_beta|  = %d at %s\n",
            length(causal_beta),  paste(causal_beta,  collapse = ", ")))
cat(sprintf("Spillover truth |S_delta| = %d at %s\n",
            length(causal_delta), paste(causal_delta, collapse = ", ")))
```

## 6. Fit a single model

If you already know your penalties, call `heritage()` directly. Both
`lambda_beta` and `lambda_delta` are required.

```{r fit-single}
fit <- heritage(y, X, sX,
                family       = "gaussian",
                lambda_beta  = 0.05,
                lambda_delta = 0.05,
                verbose      = FALSE)
print(fit)
```

`summary()` lists every non-zero coefficient:

```{r summary-single}
summary(fit)
```

## 7. Select penalties by grid search

In practice you rarely know the penalties in advance. `heritage_gs()` searches
a grid of `(lambda_beta, lambda_delta)` pairs, selecting the combination that
optimises a metric on a held-out **tuning** set, then reports performance on an
untouched **test** set.

To avoid leakage between relatives, we split **by family** rather than by
individual.

```{r split}
family_id <- paste(pedigree$dam, pedigree$sire, sep = "-")
fams      <- unique(family_id)

set.seed(42)
fams      <- sample(fams)
n_tr      <- round(0.6 * length(fams))
n_tu      <- round(0.2 * length(fams))

f_train <- fams[seq_len(n_tr)]
f_tune  <- fams[(n_tr + 1L):(n_tr + n_tu)]
f_test  <- fams[(n_tr + n_tu + 1L):length(fams)]

i_tr <- which(family_id %in% f_train)
i_tu <- which(family_id %in% f_tune)
i_te <- which(family_id %in% f_test)

c(train = length(i_tr), tune = length(i_tu), test = length(i_te))
```

Standardise the design matrices using **training-set** means and standard
deviations only:

```{r standardise}
mu <- colMeans(X[i_tr, ]); s <- apply(X[i_tr, ], 2, sd); s[s < 1e-10] <- 1
scl <- function(m) sweep(sweep(m, 2, mu, "-"), 2, s, "/")

gs <- heritage_gs(
  y_train = y[i_tr], X_train = scl(X[i_tr, ]), sX_train = scl(sX[i_tr, ]),
  y_tune  = y[i_tu], X_tune  = scl(X[i_tu, ]), sX_tune  = scl(sX[i_tu, ]),
  y_test  = y[i_te], X_test  = scl(X[i_te, ]), sX_test  = scl(sX[i_te, ]),
  family         = "gaussian",
  n_lambda_beta  = 12,
  n_lambda_delta = 12,
  metric         = "mse",
  verbose        = FALSE
)
print(gs)
```

The best refit model is stored in `gs$best_model` and behaves like any
`heritage` fit:

```{r best}
best <- gs$best_model
length(best$active_beta)   # selected direct effects
length(best$active_delta)  # selected spillover effects
```

## 8. Diagnostics and prediction

The grid-search object has a `plot()` method for regularisation paths and
sparsity:

```{r plot-paths, fig.cap = "Active coefficients along each penalty."}
plot(gs, type = "paths")
```

Predict on new data with the `predict()` method, supplying both `newX` and
`newsX`:

```{r predict}
y_hat <- predict(best,
                 newX  = scl(X[i_te, ]),
                 newsX = scl(sX[i_te, ]),
                 type  = "response")

r2 <- 1 - sum((y[i_te] - y_hat)^2) / sum((y[i_te] - mean(y[i_te]))^2)
cat(sprintf("Test R2 = %.3f\n", r2))

plot(y[i_te], y_hat,
     xlab = "Observed", ylab = "Predicted",
     main = sprintf("Test set (R2 = %.3f)", r2),
     pch = 19, col = "#2c7fb8")
abline(0, 1, col = "red", lty = 2)
```

## Recap

The typical HERITAGE pipeline is:

1. `build_ancestor_list()` -- enumerate each individual's ancestors;
2. `compute_spillover_X()` -- build the kinship-weighted spillover matrix;
3. `heritage_gs()` -- choose the penalties on a held-out tuning set
   (or `heritage()` if they are known);
4. `predict()` / `summary()` / `plot()` -- inspect and apply the fit.

## Reference

Diakite, A. O., Madore, A.-M., Greenwood, C. M. T., & Laprise, C. (2026).
HERITAGE: Hierarchical effects regression with interactions for trait analysis
in genetics. *Manuscript in preparation.*
